\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-06-25 manuscript/paper.tex
changes after Antoine's review

\paragraph{The high-dimensional longitudinal data context.}
With the exploding number of daily internet users, or with the development of electronic health records in a medical context, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

@@

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance the time when a client stops using a company's product or service in a churn prediction setting~ ; or survival time, re-hospitalization, relapse or disease progression in a medical context.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

@@

We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
  \label{eq:link-function}
  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\begin{equation}
   \label{eq:random-effect-law}
   b^l \sim \cN(0, D_{ll})
 \end{equation} 
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
  \label{eq:intensity-model}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
the history of the true longitudinal process up to time $t$.
For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.

@@

Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way. Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
\begin{table}[htb]
\centering
\begin{tabular}{cccc}
\toprule
Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$  & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Examples of classical shared associations in SREMs.}
\label{table:shared_associations}
\end{table}

@@

  \path [->,draw,ultra thick, blue]
        (G) edge (mu)
        (Y) edge (W);
  \path [->,draw,ultra thick]
        (beta) edge (mu)
        (b) edge (mu)
        (mu) edge (Y)
        (Sigma) edge (b)
        (phi) edge (Y)
        (xi) edge (G)
        (X) edge (G)
        (gamma) edge (lambda)
        (W) edge (lambda)
        (G) edge (lambda)
        (lambda) edge (T)
        ;

@@

\end{tikzpicture}}
\caption{Graphical representation of the lights model. Square boxes denote observed data while circles denote unobserved (including random) terms. One can then observe the two links in blue between the submodels that gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$ (defined in~\eqref{eq:dataset-def}), corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} S_k(t_i) , \]
and
\begin{equation}
  \label{eq:survival-k-def}
  S_k(t_i) = \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}
\end{equation}
is the survival probability of subject $i$ given that it belongs to latent class $k$.
As expected,~\eqref{eq:log-lik} is a JLCMs like log-likelihood.

@@

One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important (time-dependent and time-independent) features relatively to the prediction objective. The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership. The ridge part allows to handle potential correlation between time-independent features.
On the other hand, for the sparse group lasso penalty, a group $l$ corresponds to a trajectory, that is one longitudinal outcome. So if $\hat \gamma_k^l$ is entirely zeroed (thanks to the group lasso part), it means that the $l$-th longitudinal process is discarded by the model in terms of risk effect for the $k$-th latent class.
Then, the sparse part of the penalty allows to perform representation features selection for each trajectory : for $\hat \gamma_k^l$ that are not entirely zeroed, its support informs on the representation features involved in the $k$-th latent class risk of event for the $l$-th longitudinal outcome.
Note that in practice, the intercept terms are not regularized.

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation*}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation*}
where
\begin{equation}
  \label{eq:Lambda_ik-b_i}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} =  f_{\theta^{(w)}}(t_i, \delta_i | G_i = k) f_{\theta^{(w)}}(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | y_i, G_i = k) \dd b_i,
\end{equation}
and in the Gaussian case used in practice one has
\begin{align*}
  y_i | G_i = k ; \theta^{(w)} &\sim \cN\big(U_i\beta_k^{(w)}, V_iD^{(w)}V_i^\top + \Sigma_i^{(w)}\big), \\
  b_i | y_i, G_i=k ; \theta^{(w)} &\sim \cN \big(O_{i,k}^{(w)}, W_i^{(w)}\big),  
\end{align*}
with $O_{i,k}^{(w)}=W_i^{(w)}V_i^T{\Sigma_i^{(w)}}^{-1}(y_i - U_i\beta_k^{(w)})$ and $W_i^{(w)} = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$. Hence, what is interesting here is that~\eqref{eq:Lambda_ik-b_i} is actually closed-formed since 
\[ \E_{\theta^{(w)}}[ \tilde g_1(b_i) | y_i, G_i = k] = O_{i,k}^{(w)} \quad \text{and} \quad \E_{\theta^{(w)}}[ \tilde g_2(b_i) | y_i, G_i = k] = W_i^{(w)} + O_{i,k}^{(w)}{O_{i,k}^{(w)}}^\top\]
and then tractable analytically. Whereas in classical SREMs, some form of approximation must be used in practice to compute this quantity, like numerical integration techniques (for instance Gaussian quadrature) or Monte Carlo approximation.

@@

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:evaluation strategy}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}. However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power. So we choose to put ourselves in what we call a ``real-time prediction paradigm''.

@@

We detail in this section the chosen metric to evaluate risk prediction performances. 
Let us denote by $\hat \cR_i$ the predictive marker for a subject $i$ at $t^{max}_i$ and from a given model (for the lights, $\hat \cR_i = \hat \cR_{ik}$ in~\eqref{eq:marker-def} with $k=1$).
A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
\begin{equation*}
  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
\end{equation*}
with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
\begin{equation*}
  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
\end{equation*}
with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}. A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
Hence in the following, we consider the C-index metric to assess performances.

@@

We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model. Given the latent class membership, each submodel is assumed to be independent. In this context, the predictive marker for subject $i$ is
\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\}.\]

\texttt{@Antoine : est-ce qu'il peut prendre des time-indep feat ? et quelles sont les assos ?}

>>>>

\paragraph{The high-dimensional longitudinal data context.}
With the exploding number of daily internet users, or with the development of electronic health records in a medical context, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple joint models separately each one with univariate longitudinal outcome~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

@@

The setting of this paper is such that we want to incorporate high-dimensional trajectories, or longitudinal features, measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance the time when a client stops using a company's product or service in a churn prediction setting~ ; or survival time, re-hospitalization, relapse or disease progression in a medical context.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

@@

We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
  \label{eq:link-function}
  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\begin{equation}
   \label{eq:random-effect-law}
   b^l \sim \cN(0, D_{ll})
 \end{equation} 
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix, which is not class-specific.

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
  \label{eq:intensity-model}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function that does not depend on $k$, and we denote 
\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
the history of the true longitudinal process up to time $t$.
Then for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping (also independent from $k$), and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters, being the only class-specific object in the risk of event submodel.

@@

Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way. Table~\ref{table:shared_associations} gives common parameterizations for shared associations in SREMs. It is now common to include several of these in the survival sub-model, see for instance~\citet{}, but as far as we know, there is no method to consider simultaneously a huge number of associations and let the model learn which one is predictive for the underlying task it is trained for.

TODO : add REF :

Andrinopoulou, E.-R. and Rizopoulos, D. (2016). Bayesian shrinkage approach for a joint model of
longitudinal and survival outcomes assuming different association structures. Statistics in Medicine,
35(26):4813–4823.

Ferrer, L., Rondeau, V., Dignam, J., Pickles, T., Jacqmin-Gadda, H., \& Proust-Lima, C. (2016). Joint modelling of longitudinal and multi-state processes: application to clinical progressions in prostate cancer. Statistics in medicine, 35(22), 3933–3948. https://doi.org/10.1002/sim.6972

\begin{table}[htb]
\centering
\begin{tabular}{cccc}
\toprule
Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$  & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Examples of classical shared associations in SREMs.}
\label{table:shared_associations}
\end{table}

@@

  \path [->,draw,ultra thick, blue]
        (G) edge (mu)
        (G) edge (lambda)
        (Y) edge (W);
  \path [->,draw,ultra thick]
        (beta) edge (mu)
        (b) edge (mu)
        (mu) edge (Y)
        (Sigma) edge (b)
        (phi) edge (Y)
        (xi) edge (G)
        (X) edge (G)
        (gamma) edge (lambda)
        (W) edge (lambda)
        (lambda) edge (T)
        ;

@@

\end{tikzpicture}}
\caption{Graphical representation of the lights model, with $\phi_l$ being defined in~\eqref{eq:def-Phi_i} below. Square boxes denote observed data while circles denote unobserved (including random) terms. One can then observe the two links in blue between the submodels that gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$ (defined in~\eqref{eq:dataset-def}), corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} S_k(t_i) , \]
and
\begin{equation}
  \label{eq:survival-k-def}
  S_k(t_i) = \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}
\end{equation}
is the survival probability of subject $i$ given that it belongs to latent class $k$.
As expected,~\eqref{eq:log-lik} is a JLCMs like log-likelihood. Note that to write~\eqref{eq:log-lik}, one could simplify Assumption~\ref{indep-hyp-3} by removing the $b_i$ both in the left and right side of Equation~\eqref{eq:ind-hyp} to get a classical JLCMs hypothesis, but the $b_i$ mention is useful later on in the E-step of the inference algorithm, for instance to compute Equation~\eqref{eq:E_g}.

@@

One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important features (trajectories and time-independent) relatively to the prediction objective. The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership. The ridge part allows to handle potential correlation between time-independent features.
On the other hand, for the sparse group lasso penalty, a group $l$ corresponds to a trajectory, that is one longitudinal outcome. So if $\hat \gamma_k^l$ is entirely zeroed (thanks to the group lasso part), it means that the $l$-th longitudinal process is discarded by the model in terms of risk effect for the $k$-th latent class.
Then, the sparse part of the penalty allows to perform representation features selection for each trajectory : for $\hat \gamma_k^l$ that are not entirely zeroed, its support informs on the representation features involved in the $k$-th latent class risk of event for the $l$-th longitudinal outcome.
Note that in practice, the intercept terms are not regularized.

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1^\top, \ldots, b_n^\top)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation}
  \label{eq:E_g}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
where
\begin{equation}
  \label{eq:Lambda_ik-b_i}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} =  f_{\theta^{(w)}}(t_i, \delta_i | G_i = k) f_{\theta^{(w)}}(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | y_i, G_i = k) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$. 
In the Gaussian case used in practice one has
\begin{align*}
  y_i | G_i = k ; \theta^{(w)} &\sim \cN\big(U_i\beta_k^{(w)}, V_iD^{(w)}V_i^\top + \Sigma_i^{(w)}\big), \\
  b_i | y_i, G_i=k ; \theta^{(w)} &\sim \cN \big(O_{i,k}^{(w)}, W_i^{(w)}\big),  
\end{align*}
with $O_{i,k}^{(w)}=W_i^{(w)}V_i^T{\Sigma_i^{(w)}}^{-1}(y_i - U_i\beta_k^{(w)})$ and $W_i^{(w)} = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$. Hence, what is interesting here is that~\eqref{eq:Lambda_ik-b_i} is actually closed-formed since 
\[ \E_{\theta^{(w)}}[ \tilde g_1(b_i) | y_i, G_i = k] = O_{i,k}^{(w)} \quad \text{and} \quad \E_{\theta^{(w)}}[ \tilde g_2(b_i) | y_i, G_i = k] = W_i^{(w)} + O_{i,k}^{(w)}{O_{i,k}^{(w)}}^\top\]
and then tractable analytically. Whereas in classical SREMs, some form of approximation must be used in practice to compute this quantity, like numerical integration techniques (for instance Gaussian quadrature) or Monte Carlo approximation.

@@

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in~\eqref{eq:c-index-def}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}. However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power. So we choose to put ourselves in what we call a ``real-time prediction paradigm''.

@@

We detail in this section the chosen metric to evaluate risk prediction performances. 
Let us denote by $\hat \cR_i$ the predictive marker for a subject $i$ at $t^{max}_i$ and from a given model (for the lights, $\hat \cR_i = \hat \cR_{ik}$ in~\eqref{eq:marker-def} with $k=1$).
A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
\begin{equation*}
  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
\end{equation*}
with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
\begin{equation}
  \label{eq:c-index-def}
  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
\end{equation}
with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}. A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
Hence in the following, we consider the C-index metric to assess performances.

@@

We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
There is no shared associations between the longitudinal models and the survival model. Given the latent class membership, each submodel is assumed to be independent. In this context, the predictive marker for subject $i$ is
\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\},\] 
where the shared associations $\varphi_{k,a}$ are the four given in Table~\ref{table:shared_associations}.
